From Reversible Programs to \\ Univalent Universes and Back

### 2017-06-30 upi.tex
remove some old material which was entirely subsumed and no longer served any purpose.

%% \jacques{Example 1: using the combinators from pi-dual, some an equational
%% derivation in type theory (which?), in equational-derivation style.  Point out
%% that reading ``down'' the justifications is a $\Pi$ program which is also a
%% constructive proof of the equivalence of the types.  As it is reversible, we
%% can also read the reverse program by reading ``up''.}

%% \jacques{Refer to some Sabry et al. papers regarding Toffoli gates.}

% \begin{comment}
% In this paper we will
% identify a ``featherweight'' representative to use for the formal
% development. But first, we illustrate the flavor of the family of languages with
% some examples written in a Haskell syntax. First consider the following program:

% {\footnotesize
% \begin{verbatim}
% addSub1 :: (Int, Int) :<=> (Int, Int)
% addSub1 = CommuteTimes
%       :.: (UnfoldN :*: Id)
%       :.: Distribute
%       :.: (Id :+: CommuteTimes)
%       :.: Factor
%       :.: (FoldN :*: Id)
% \end{verbatim}}
% %
% \noindent which takes a pair of natural numbers and processes them as follows depending on
% whether the second number is 0 or not. If the second number is 0, the evaluation
% proceeds as follows:
% $(n,0) \mapsto (0,n) \mapsto (\mathit{inj}_1(),n) \mapsto \mathit{inj}_1((),n)
% \mapsto \mathit{inj}_1((),n) \mapsto (\mathit{inj}_1(),n) \mapsto (0,n)$. If the
% second number is $\mathit{suc}~m$, the evaluation proceeds as follows:
% $(n, \mathit{suc}~m) \mapsto (\mathit{suc}~m,n) \mapsto (\mathit{inj}_2m,n)
% \mapsto \mathit{inj}_2(m,n) \mapsto \mathit{inj}_2(n,m) \mapsto
% (\mathit{inj}_2n,m) \mapsto (\mathit{suc}~n,m)$. In summary, repeated
% applications of this program would increment the first number and decrement the
% second number maintaining the sum at all times. When the second number reaches
% 0, the numbers are swapped.

% As another example, consider the following expressions culminating with the
% implementation of the Toffoli gate, which is universal for combinational
% circuits. We have two implementations of simple boolean negation: a direct one
% and a more convoluted one. These two implementations are ``equivalent'' in the
% sense to be explained below. The expression \verb|cond| is the main reversible
% control flow mechanism: it allows for the conditional execution of one of two
% branches but retains the choice of which branch was executed. This mechanism can
% be used twice to implement the Toffoli gate, which performs a conditional
% negation on one bit if two other bits are both true:

% {\footnotesize
% \begin{verbatim}
% inot :: Bool :<=> Bool
% inot = CommutePlus

% inot2 :: Bool :<=> Bool
% inot2 = UnitTimesI
%     :.: CommuteTimes
%     :.: (CommutePlus :*: Id)
%     :.: CommuteTimes
%     :.: UnitTimesE

% cond :: (a :<=> b) -> (a :<=> b) -> ((Bool, a) :<=> (Bool, b))
% cond f g = Distribute
%        :.: ((Id :*: f) :+: (Id :*: g))
%        :.: Factor

% controlled :: (a :<=> a) -> ((Bool, a) :<=> (Bool, a))
% controlled f = cond f Id

% cnot :: (Bool, Bool) :<=> (Bool, Bool)
% cnot = controlled inot

% toffoli :: ((Bool,Bool),Bool) :<=> ((Bool,Bool),Bool)
% toffoli = AssocTimesR :.: controlled cnot :.: AssocTimesL
% \end{verbatim}}

% Although writing circuits using the raw syntax for combinators above is tedious,
% the examples illustrate the programming language nature of this family of $\Pi$
% languages. In other work, one can find a compiler from a conventional functional
% language to circuits~\cite{James:2012:IE:2103656.2103667}, a systematic
% technique to translate abstract machines to $\Pi$~\cite{rc2012}, and a
% Haskell-like surface language~\cite{theseus} which can ease writing
% programs. All that reinforces the claim that this is a practical approach to
% reversible programming.

% The raw syntax for combinators has, however, a significant benefit: it allows
% algebraic manipulation of the programs using a second level of combinators. As
% an example it is possible to show the the two implementations of boolean
% negation are equivalent as follows:

% {\footnotesize
% \begin{verbatim}
% negOpt :: inot2 :<===> inot
% negOpt = (Id :>: AssocLeftEquiv)
%     :..: (Id :>: (SlideCommuteTimes :>: Id))
%     :..: (Id :>: AssocRightEquiv)
%     :..: (Id :>: (Id :>: AssocLeftEquiv))
%     :..: (Id :>: (Id :>: (InvInvEquivId :>: Id)))
%     :..: (Id :>: (Id :>: IdElimL))
%     :..: AssocLeftEquiv
%     :..: (SlideUnitI :>: Id)
%     :..: AssocRightEquiv
%     :..: (Id :>: InvInvEquivId)
%     :..: IdElimR
% \end{verbatim}}
% %
% \noindent This program, which can be thought of as an optimization or a proof,
% relates the two implementations of boolean negation via a series of elementary
% optimizations of which the following are the most interesting:

% {\footnotesize
% \begin{verbatim}
% AssocLeftEquiv    ::           (c1 :.: (c2 :.: c3)) :<===> ((c1 :.: c2) :.: c3)
% SlideCommuteTimes :: (CommuteTimes :.: (c1 :*: c2)) :<===> ((c2 :*: c1) :.: CommuteTimes)
% IdElimL           ::                     (Id :.: c) :<===> c
% SlideUnitI        ::   (UnitTimesI :.: (c1 :*: c2)) :<===> (c2 :.: UnitTimesI)
% \end{verbatim}}

% \end{comment}

\jacques{Example 2: like example 1, but this time at the next level (i.e. a
program equivalence).  Same remark old.  Refer to Shulmann notes, and point out
that such equational proofs at this level can actually be interpreted in appropriate
categories as commutative diagrams.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
~
\newpage
\section{Old}

The first step was a refinement and extension of the
language $\Pi$ with a layer of program transformations between the programs
witnessing type isomorphisms~\cite{Carette2016}. The second crucial result was
the formal presention

Second,

POPL 2012: Information effects; Canonicity for 2-dimensional type theory; look
different but we show are the ``same''.

There is a cottage industry of reversible programming languages, reversible
logic, programming applications of type isomorphisms, etc. It seems that this
work should be connected to HoTT and univalence, but there aren't any precise
connections or theorems.

We propose a very specialized version of HoTT, which we dub
``featherweight HoTT''; it sports a smaller
collection of types and only reversible functions. No dependent types!  More
precisely, our conjecture:

\begin{itemize}
\item Let $\mathcal{U}$ be the univalent subuniverse generated by $0$, $1$, $+$,
  and $*$ suitably truncated.
\item Let $\Pi$ be the previously  described reversible programming with 1-paths and 2-paths
\item We conjecture that $\Pi$ includes codes for \emph{all} paths in $\mathcal{U}$
\end{itemize}

\jacques{Just to make sure: we're aiming to prove a much much smaller version of
this here, right?  Where in fact the only type is Bool?}

 Following the HoTT book and the slides on ``A Characterization of Univalent
Fibrations'' by Dan Christensen. All the definitions and most of the examples in
this section are formalized in the accompanying Agda code.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\section{Old}
